---
title: Easy Tables from Statistical Output (web version)
author: "Barry DeCicco"
date: "For Ann Arbor R Users' Group Meetup, May 11, 2023"
output:  
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding:  hide
    

---


```{r load-libraries, include=FALSE}

# load libraries as needed:

library(tidyverse)
library(janitor)
library(kableExtra)
library(magrittr)
library(tidyverse)
library(summarytools)
library(xtable)
library(broom)
library(tidyr)
library(foreign)
library(haven)
library(here)

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

here::i_am(path = "inst/Easy Tables from Statistical Output - HTML Version - Updated.Rmd")

```



# Making tables from Statistical Output

Copying and pasting statistical output is time-consuming
and prone to errors.  When multiple analyses need to be run,
it's even worse.

R has multiple ways to extract organized statistical ouput into data frames or tables. At that point the information is easily understood 
and renderable in documents and presentations.

 This presentation will cover {.flexbox .vcenter}

*   The nature of R statistical output and how to explore it
and extract items 'by hand'.

*   How to use the **xtable** pacakge to extract and organize output.<p>

*   How to use the **broom** package to clean and tidy output.<p>

*   How to make tables look better with the kable and KableExtra packages.<p>



# Data - bodyfat.sav:

*   This presentation will use the 'BodyFat' data set built used as a sample set for SPSS (available at http://calcnet.mth.cmich.edu/org/spss/prjs_datasets.htm )



```{r load-bodyfat-data-set, echo = TRUE}

bodyfat <- read_sav(here("inst/extdata/bodyfat.sav"))

```

## Body Fat Data Set:

```{r print-bodyfat-data-set, echo = TRUE}

print(bodyfat_df)

```


# Basic Structure of Output


 - Assign the output of a model to an object.
 - Use 'class()' to find out the class of the output object (you can then google it)
 - Use the 'names()' function to find the names of the major components of the object.
 - Use the 'str()' to get a breakdown of the object's components

## Simple Linear Model

Typing the object's name gives some basic information.

```{r run-lm, echo =TRUE}

lm1 <- lm(y ~ x1, data= bodyfat_df)

```


```{r print-lm, echo =TRUE}

lm1 


```



```{r class-lm, echo =TRUE}

class(lm1)

```


```{r names-lm, echo =TRUE}

names(lm1)

```


```{r str-lm, echo =TRUE}

str(lm1)

```


You can extract the parts of the object, examine them and then use them:

```{r print-coefficients-lm}

lm1$coefficients

```


```{r structure-coefficients-lm}

names(lm1$coefficients)
str(lm1$coefficients)


```


```{r extract-coefficients-lm}

class(lm1$coefficients)
dim(lm1$coefficients)
length(lm1$coefficients)
lm1$coefficients[1]
lm1$coefficients[2]

```


You can also use the summary - this will print out different information
depending on the class of the object:


```{r summary-lm, echo =TRUE}

summary(lm1) 

```


By examining the structure of the summary you can get another view of the object:


```{r str-summary-lm, echo =TRUE}

str(summary(lm1) )

```
At this point you can attempt to access the coefficients *and* the std. errors:

```{r coefficients-standard-errors-lm1, echo =TRUE}

summary_coefficients <- summary(lm1)$coefficients
class(summary_coefficients)

```

Once you can refer to these values, you can now use these values in text:

```{r use-values-in-text, echo =TRUE}

# The estimated effect of X1 is: `r round(summary_coefficients[2,1],2)`; 
# standard error is: `r round(summary_coefficients[2,2],2)`.

```

The estimated effect of X1 is: `r round(summary_coefficients[2,1],2)`; 
standard error is: `r round(summary_coefficients[2,2],2)`.

And so on.  

You could also run chunks through kable, to get a nice looking table for use in 
papers or presentations:

```{r, echo =TRUE}

summary_coefficients %>% 
  as.data.frame() %>% 
  dplyr::mutate(across(where(is.numeric)), round) %>% 
  kable() %>% 
kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) 


```






# What Else Can You Do with the Results?



```

# Using Kable with Other Packages.

Some other packages have an '...2kable() function to allow you to use kable:

xtable2able()
For the 'table' package, use toKable()

## Other things: 

- Formatting headers (e.g., a higher row covering groups of variables)
- Controlling the cells (e.g., conditional highlighting).
- Linking to other locations.





# Questions?

# References:

 - The janitor package vignette:  https://garthtarr.github.io/meatR/janitor.html
 
 - The kableExtra package:  https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
 


    

