---
title: "Making Tables Easy with Tabyl and kableExtra (Web Version)"
author: "Barry DeCicco"
date: "For ABS Network Presentation, May 3, 2023"
output:  
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding:  hide
    

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r load-libraries, include=FALSE}

# load libraries as needed:

library(tidyverse)
library(janitor)
library(kableExtra)
library(magrittr)
library(tidyverse)
library(summarytools)
```


# This presentation will cover {.flexbox .vcenter}


*   How to make summary tables with the tabyl function of the janitor package.<p>
*   How to make summary tables with the describe function of the Hmisc/psych packages.<p>
*   How to make sumamry tables with the summaryM function of the Hmisc package.<p>
*   How to make tables look better with the kable and KableExtra packages.<p>


# Janitor and the Tabyl function:

*   The janitor package has many cleaning functions, plus the tabyl fuction,
for making tables.<p>
*   Those tables can be enhanced with other functions in the janitor package.<p>
*   they can be converted to data frames or tibbles, and then manipulated
with other R commands, such as dplyr.<p>
*   Data frames can be rendered by the kable and kableExtra packages, 
for better printing.<p>



# Data - mtcars:

*   This presentation will use the 'mtcars' data set built into base R, 
and a subset (first 5 rows / first 4 columns) which will be called 
'mtcars_trimmed'.
*   This will then be enhanced.

```{r cars_print_1, echo = TRUE}

head(mtcars)

```

## mtcars_trimmed:

```{r cars_print_2, echo = TRUE}
mtcars_trimmed <- mtcars[1:5, 1:4]
print(mtcars_trimmed)

```


# The janitor package

## The janitor does 'clean up' tasks:

 - Perfectly format data frame column names;
 - Isolate partially-duplicate records; and
 - Provide quick tabulations (i.e., frequency tables and crosstabs).

## For now, just look at the tabulations, which are done with 'tabyl()'



## Basic Table:



```{r tabyl_1, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE)




```


## Adding to the Basic Tables, using 'Adorn':

  * The Adorn commands can be used on data frames
  
  * The output of the tabyl() command can be converted to a data frame using
  the command as.data.frame()


```{r tabyl_2, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE) %>% 
  adorn_totals(where="row")




```

## Using 'Adorn' with multiple options:



```{r tabyl_3, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE) %>% 
  adorn_totals(where=c("row","col"))




```


## The use of Adorn can be extended:



```{r tabyl_4, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE)       %>% 
  adorn_totals(where=c("row","col"))     %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_pct_formatting(digits = 0) 




```


## Mixing in Percentages, and Formatting Them:


```{r tabyl_5, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE)       %>% 
  adorn_totals(where=c("row","col"))     %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")




```



# The kableExtra package

## This package allows easy html/latex/rtf tables (we will stick with html here).

Here is the simplest, with no options:


```{r kableExtra_1, echo =TRUE}



mtcars %>% 
  kable()




```



## The easiest improvement is to use the styling command: 



```{r kableExtra_2, echo =TRUE}



mtcars                   %>% 
  kable()            %>%
  kable_styling()




```



## Options for the styling() commands: 

- The 'responsive' command will allow horizontal scrolling on small screens;
it does not show up on wider screens.
- The 'full_width = F' will not take up the whole page width if not needed.
- The position = 'center' command centers it ('left', 'right')
- (not shown) "float_right" would allow text to be put to the left of the table.

## Code:



```{r kableExtra_3, echo =TRUE}

mtcars %>% 
  kable()              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                )


```

## Formatting Columns:

The unadjusted printing is bad, and could be improved:


```{r kableExtra_4, echo =TRUE}

mtcars                     %>% 
  kable()              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "2cm") 

```



## Locking the Header

This locks the header at the top of the screen (not visible for small tables)

```{r kableExtra_5, echo =TRUE}

mtcars                     %>% 
  kable()              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "2cm",bold = TRUE) %>% 
  kable_styling(fixed_thead = T)

```


## Column Headers

You can substitute your columns names for the names in the data frame.

**It's important to revise those names as you revise the table**

'escape = FALSE' is an option which renders '<br/>' as a line return.

```{r kableExtra_6, echo =TRUE, eval=TRUE}

mtcars            %>% 
  kable(escape = FALSE,
        col.names = c(
                      "Miles per<br/>Gallon",
                      "Cylinders",
                      "Displacement",
                      "Horsepower",
                      "Rear Axle<br/>Ratio",
                      "Weight",
                      "Quarter Mile<br/>
                      Seconds",
                      "V vs Straight<br/>Line",
                      "Automatic vs <br/> Manual Transmission",
                      "Gear",
                      "Carburetors"
                      ))              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "3cm") %>% 
  kable_styling(fixed_thead = T)
  



```

## Titles and Footnotes

You can set up both in separate statements, which give you more flexibility.
If you parametrize them, then you can have more automatric updating

```{r kableExtra_7, echo =TRUE, eval=TRUE}

 "`r format(Sys.time(), '%d %B, %Y')`"

title_date = format(Sys.time(), '%d %B, %Y')

T1cap <- paste0("<center><strong>TITLE TEXT GOES HERE,<br/>", title_date, "</strong></center>>")
FN1   <- paste("<strong>FOOTNOTE TEXT GOES HERE/<strong>")
mtcars            %>% 
  kable(caption = T1cap, 
        escape = FALSE,
        col.names = c(
                      "Miles per<br/>Gallon",
                      "Cylinders",
                      "Displacement",
                      "Horsepower",
                      "Rear Axle<br/>Ratio",
                      "Weight",
                      "Quarter Mile<br/>
                      Seconds",
                      "V vs Straight<br/>Line",
                      "Automatic vs <br/> Manual Transmission",
                      "Gear",
                      "Carburetors"
                      ))              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "3cm") %>% 
  kable_styling(fixed_thead = T)  %>% 
  footnote(general = FN1, general_title = "Note: ", footnote_as_chunk = TRUE)
 



```



## Header Rows

You can put header rows, including groups of columns.


```{r kableExtra_8, echo =TRUE, eval=TRUE}

 "`r format(Sys.time(), '%d %B, %Y')`"

title_date = format(Sys.time(), '%d %B, %Y')

T1cap <- paste0("<center><strong>TITLE TEXT GOES HERE,<br/>", title_date, "</strong></center>>")
FN1   <- paste("<strong>FOOTNOTE TEXT GOES HERE/<strong>")
mtcars            %>% 
  kable(caption = T1cap, 
        escape = FALSE,
        col.names = c(
                      "Miles per<br/>Gallon",
                      "Cylinders",
                      "Displacement",
                      "Horsepower",
                      "Rear Axle<br/>Ratio",
                      "Weight",
                      "Quarter Mile<br/>
                      Seconds",
                      "V vs Straight<br/>Line",
                      "Automatic vs <br/> Manual Transmission",
                      "Gear",
                      "Carburetors"
                      ))              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "3cm") %>% 
  kable_styling(fixed_thead = T)  %>% 
  add_header_above(c(" " = 2, "Middle 4 Columns" = 4, 
                     "Last 5 Columns" = 5, " ")) %>% 
  add_header_above(c(" " = 4, "Group 1" = 5, " " = 3)) %>%
  add_header_above(c(" "=1,"Everything" = 11)) %>% 
  footnote(general = FN1, general_title = "Note: ", footnote_as_chunk = TRUE)
 



```


## Putting the Table into a Scrollbox:

Use the command:  scroll_box(width = "500px", height = "200px")

This creates a scroll box 500x200 pixels.

You can also specifiy the width with a percentage, e.g., 
scroll_box(width = "100%", height = "200px")

```{r kableExtra_9, echo =TRUE, eval=TRUE}


title_date = format(Sys.time(), '%d %B, %Y')

T1cap <- paste0("<center><strong>TITLE TEXT GOES HERE,<br/>", title_date, "</strong></center>>")
FN1   <- paste("<strong>FOOTNOTE TEXT GOES HERE/<strong>")
mtcars            %>% 
  kable(caption = T1cap, 
        escape = FALSE,
        col.names = c(
                      "Miles per<br/>Gallon",
                      "Cylinders",
                      "Displacement",
                      "Horsepower",
                      "Rear Axle<br/>Ratio",
                      "Weight",
                      "Quarter Mile<br/>
                      Seconds",
                      "V vs Straight<br/>Line",
                      "Automatic vs <br/> Manual Transmission",
                      "Gear",
                      "Carburetors"
                      ))              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "3cm") %>% 
  kable_styling(fixed_thead = T)  %>% 
  add_header_above(c(" " = 2, "Middle 4 Columns" = 4, 
                     "Last 5 Columns" = 5, " ")) %>% 
  add_header_above(c(" " = 4, "Group 1" = 5, " " = 3)) %>%
  add_header_above(c(" "=1,"Everything" = 11)) %>% 
  footnote(general = FN1, general_title = "Note: ", footnote_as_chunk = TRUE) %>% 
  scroll_box(width = "75%", height = "500px")
 



```

# Using the Table Elsewhere.


## Putting the Table into a Scrollbox:

Use the command:  scroll_box(width = "500px", height = "200px")

This creates a scroll box 500x200 pixels.

You can also specifiy the width with a percentage, e.g., 
scroll_box(width = "100%", height = "200px")

```{r kableExtra_10, echo =TRUE, eval=TRUE}


title_date = format(Sys.time(), '%d %B, %Y')

T1cap <- paste0("<center><strong>TITLE TEXT GOES HERE,<br/>", title_date, "</strong></center>>")
FN1   <- paste("<strong>FOOTNOTE TEXT GOES HERE/<strong>")
mtcars            %>% 
  kable(caption = T1cap, 
        escape = FALSE,
        col.names = c(
                      "Miles per<br/>Gallon",
                      "Cylinders",
                      "Displacement",
                      "Horsepower",
                      "Rear Axle<br/>Ratio",
                      "Weight",
                      "Quarter Mile<br/>
                      Seconds",
                      "V vs Straight<br/>Line",
                      "Automatic vs <br/> Manual Transmission",
                      "Gear",
                      "Carburetors"
                      ))              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "3cm") %>% 
  kable_styling(fixed_thead = T)  %>% 
  add_header_above(c(" " = 2, "Middle 4 Columns" = 4, 
                     "Last 5 Columns" = 5, " ")) %>% 
  add_header_above(c(" " = 4, "Group 1" = 5, " " = 3)) %>%
  add_header_above(c(" "=1,"Everything" = 11)) %>% 
  footnote(general = FN1, general_title = "Note: ", footnote_as_chunk = TRUE) %>% 
  scroll_box(width = "75%", height = "500px")
 



```



# What Else Can You Do with the Results?

You can use the command '' to export the table to a file, if you would like to
use it separately:

save_kable(file = "table1.html", self_contained = T)


```{r kableExtra_11, echo =TRUE, eval=TRUE}


title_date = format(Sys.time(), '%d %B, %Y')

T1cap <- paste0("<center><strong>TITLE TEXT GOES HERE,<br/>", title_date, "</strong></center>>")
FN1   <- paste("<strong>FOOTNOTE TEXT GOES HERE</strong>")
mtcars            %>% 
  kable(caption = T1cap, 
        escape = FALSE,
        col.names = c(
                      "Miles per<br/>Gallon",
                      "Cylinders",
                      "Displacement",
                      "Horsepower",
                      "Rear Axle<br/>Ratio",
                      "Weight",
                      "Quarter Mile<br/>
                      Seconds",
                      "V vs Straight<br/>Line",
                      "Automatic vs <br/> Manual Transmission",
                      "Gear",
                      "Carburetors"
                      ))              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                ) %>% 
  column_spec(1, width = "3cm",bold = TRUE) %>%   
  column_spec(2:5, width = "3cm") %>% 
  kable_styling(fixed_thead = T)  %>% 
  add_header_above(c(" " = 2, "Middle 4 Columns" = 4, 
                     "Last 5 Columns" = 5, " ")) %>% 
  add_header_above(c(" " = 4, "Group 1" = 5, " " = 3)) %>%
  add_header_above(c(" "=1,"Everything" = 11)) %>% 
  footnote(general = FN1, general_title = "Note: ", footnote_as_chunk = TRUE) %>% 
  scroll_box(width = "75%", height = "500px") %>% 
  save_kable(file = "table1.html", self_contained = T)
 



```

# Using Kable with Other Packages.

Some other packages have an '...2kable() function to allow you to use kable:

xtable2able()
For the 'table' package, use toKable()

## Other things: 

- Formatting headers (e.g., a higher row covering groups of variables)
- Controlling the cells (e.g., conditional highlighting).
- Linking to other locations.





# Questions?

# References:

 - The janitor package vignette:  https://garthtarr.github.io/meatR/janitor.html
 
 - The kableExtra package:  https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
 


    

