---
title: "Making Tables Easy with Tabyl and kableExtra"
author: "Barry DeCicco"
date: "For Ann Arbor R Users' Group, Feb 9, 2023"
output:
  ioslides_presentation:
    template: quarterly-report.html

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r load-libraries, include=FALSE}

# load libraries as needed:

library(tidyverse)
library(janitor)
library(kableExtra)
library(magrittr)
library(tidyverse)


```


## This presentation will cover {.flexbox .vcenter}


*   How to make summary tables with the tabyl function of the janitor package.<p>
*   How to make summary tables with the describe function of the Hmisc/psych packages.<p>
*   How to make sumamry tables with the summaryM function of the Hmisc package.<p>
*   How to make tables look better with the kable and KableExtra packages.<p>


## Janitor and the Tabyl function:

*   The janitor package has many cleaning functions, plus the tabyl fuction,
for making tables.<p>
*   Those tables can be enhanced with other functions in the janitor package.<p>
*   they can be converted to data frames or tibbles, and then manipulated
with other R commands, such as dplyr.<p>
*   Data frames can be rendered by the kable and kableExtra packages, 
for better printing.<p>



## Data - mtcars:

*   This presentation will use the 'mtcars' data set built into base R, 
and a subset (first 5 rows / first 4 columns) which will be called 'dt'.
*   This will then be enhanced.

```{r cars_print_1, echo = TRUE}

head(mtcars)

```

## dt:

```{r cars_print_2, echo = TRUE}
dt <- mtcars[1:5, 1:4]
print(dt)

```


## The janitor package

### The janitor does 'clean up' tasks:

 - Perfectly format data frame column names;
 - Isolate partially-duplicate records; and
 - Provide quick tabulations (i.e., frequency tables and crosstabs).

### For now, just look at the tabulations, which are done with 'tabyl()'



## Basic Table:



```{r tabyl_1, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE)




```


## Adding to the Basic Tables, using 'Adorn':



```{r tabyl_2, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE) %>% 
  adorn_totals(where="row")




```

## Using 'Adorn' with multiple options:



```{r tabyl_3, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE) %>% 
  adorn_totals(where=c("row","col"))




```


## The use of Adorn can be extended:



```{r tabyl_4, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE)       %>% 
  adorn_totals(where=c("row","col"))     %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_pct_formatting(digits = 0) 




```


## Mixing in Percentages, and Formatting Them:


```{r tabyl_5, echo =TRUE}



mtcars %>% 
  tabyl(gear, cyl, show_NA = TRUE)       %>% 
  adorn_totals(where=c("row","col"))     %>% 
  adorn_percentages(denominator = "col") %>% 
  adorn_pct_formatting(digits = 0) %>% 
  adorn_ns(position = "front")




```



# The kableExtra package

## This package allows easy html/latex/rtf tables (we will stick with html here).

Here is the simplest, with no options:


```{r kableExtra_1, echo =TRUE}



dt %>% 
  kable()




```



## The easiest improvement is to use the styling command: 



```{r kableExtra_2, echo =TRUE}



dt                   %>% 
  kable()            %>%
  kable_styling()




```



## Options for the styling() commands: 

- The 'responsive' command will allow horizontal scrolling on small screens.
- The 'full_width = F' will not take up the whole page width if not neeeded.
- The position = 'center' command centers it ('left', 'right')
- (not shown) "float_right" would allow text to be put to the left of the table.

## Code:



```{r kableExtra_3, echo =TRUE}

dt                     %>% 
  kable()              %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive",
                    full_width = F, position = "center")
                )


```


## Other things: 

- Locking the header row.
- Setting up scrolling.
- Controlling the cells (e.g., conditional highlighting).
- Linking to other locations.


## Hmisc Package for Summarizing Data:

The summarize function in R can be slow for many variables. The 'summaryM' command in the Hmisc package can do a lot of comparisons quickly, producing a nice table.

## Running the summary:

```{r summaryM, echo =TRUE, warning=FALSE, include=TRUE}

s <- Hmisc::summaryM( 
  mpg + disp + hp + drat + wt + qsec + vs + am + gear + carb
  ~ cyl 
  ,data=mtcars,
  overall = TRUE,
  test = TRUE
)

```


## Printing the summary:

``````{r print-summaryM, echo =TRUE, warning=FALSE, include=TRUE}

print_summary <- Hmisc::html(s,
            caption = 'Summary of Mtcars data set' , 
            exclude1,
            npct = 'both',
            digits = 3,
            prmsd = TRUE,
            brmsd = TRUE,
            mdsize = mus$maller2
            )

htmltools::save_html(print_summary, "print_summary.html")


```

## Image of the summary:


```{r include-print_summary, child='print_summary.html', echo =TRUE, warning=FALSE, include=TRUE}

```


## Questions?

## References:

 - The janitor package vignette:  https://garthtarr.github.io/meatR/janitor.html
 
 - The kableExtra package:  https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#html_only_features
 
 - The Hmisc package: 
    - Hmisc: https://www.rdocumentation.org/packages/Hmisc/versions/4.7-2
    - summaryM: Summarize Mixed Data Types vs. Groups:
    https://www.rdocumentation.org/packages/Hmisc/versions/4.7-2/topics/summaryM
    
    

